Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-23 paper/reviewer_response_PNAS.tex

\response{We have validated that one of these mutations in the transmembrane domain, C199(HA2)K, reaches reverse-genetics titers comparable to wildtype.
These data are in the new Figure S7.
In addition, various other studies have found that chimeric HAs with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB).

We posit that the disparity between the conservation of the transmembrane domain in nature and its apparent mutational tolerance in our experiments could arise at least in part from a lack of immune pressure for mutations in nature, as discussed in the response to the previous point.
}

4) Were sequences from passaged viruses (isolated on MDCKs and other cell lines) included in the comparison between in vitro mutagenesis data and natural evolution data? They need to be excluded since their presence might induce a bias towards what is tolerated/preferred in cell culture. Only sequences obtained directly from patient specimens should be used for the comparison.

\response{
This is an excellent question.
We now a separate analysis of the relationship between our measured mutational effects and maximum frequency in nature for mutations from unpassaged isolates only.
Because sequences from unpassaged isolates have only become common relatively recently, we could perform this analysis only for the post-Perth/2009 partition of the H3N2 HA phylogenetic tree.
The new Figure 5 shows that the correlation between our measurements and natural evolution is actually \emph{stronger} when we look only at unpassaged isolates.
This makes sense---lab passage mutations will not do well in nature, but will be scored favorably in our experiments.
Therefore, it makes sense that our approach works \emph{even better} when the analysis is restricted to unpassaged isolates.
We thank the reviewer for this suggestion, which has strengthened our work.
}

### 2018-06-23 paper/manuscript.tex
reviewer 1 point 4, lab passaging

We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
To calculate a given mutation's effect, we simply took the logarithm of the preference for the new amino-acid identity divided by the preference for the original amino-acid identity.
To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$).

To restrict potential biases of our experimental measurements towards mutations from cell- or egg-passaged isolates, we evaluated mutations from unpassaged isolates.
Due to the limited availability of primary isolate sequences until the last decade, we analyzed mutations of unpassaged isolates from the post-Perth partition of the phylogenetic tree~\cite{mcwhite2016sequence}.
The correlation between mutational effect and maximum frequency was stronger for mutations from unpassaged post-Perth isolates ($\rho = 0.24$).
The similar effect sizes for both partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.

>>>>

We next sought to quantify the correlation between a mutation's experimentally measured effect and the maximum frequency it attained during natural evolution.
To calculate a given mutation's effect, we simply took the logarithm of the ratio of the preferences for the mutant and wildtype amino acids at that site.
To minimize effects related to the genetic background of the strain used in the experiment, we excluded mutations closely related to the experimental strain itself and partitioned the remaining mutations into 1,022 mutations pre-dating and 299 mutations post-dating the Perth/2009 strain (Figure~\ref{suppfig:tree}).
We additionally excluded mutations from the post-Perth partition that were sampled in 2014 or after that have not had enough time for their evolutionary fates to be fully resolved.
We used these pre-Perth and post-Perth partitions to test the utility of our measurements for both post-hoc and prospective analyses, respectively.
We quantified the relationship between mutational effects and maximum mutation frequencies in the H3N2 phylogeny via Spearman rank correlation (Figure~\ref{fig:muteffect_maxfreq}A).
In both pre-Perth and post-Perth time periods, we found a modest, but statistically significant relationship between mutational effect and maximum mutation frequency (pre-Perth $\rho = 0.17$, post-Perth $\rho = 0.14$). 
The similar effect sizes for both the pre- and post-Perth partitions shows that our experimental measurements can help explain the evolutionary fates of mutations in strains that post-date the experimentally studied strain, as well as to retrospectively analyze mutations that precede the experimental strain.

Many of the HAs in sequence databases were obtained from virus isolates that were passaged in cell-culture or eggs, which can lead to the acquisition of lab-adaptation mutations that can confound evolutionary analyses~\cite{mcwhite2016sequence}.
To check that our results were not confounded by such lab-adaptation mutations, we repeated our analysis using only HA sequences that were annotated as being derived from viruses that had not been passaged in the lab.
Because sequencing of unpassaged primary isolates has only become commonplace relatively recently, we were only able to perform this analysis for the post-Perth partition of the phylogenetic tree.
Figure~\ref{fig:muteffect_maxfreq}A shows that the correlation between our measured mutational effects and the maximum frequency was even stronger for mutations from unpassaged viral isolates ($\rho = 0.24$).
